Formal Languages and Automatas

### 2024-08-11 paper/main.tex
done with background

we can always just number a finite amount of objects. The above definition

@@

  $\Sigma$. The empty words which contains no characters, 

@@

M is a set of words and plus is concetanation of words, then 

@@

@@

    \end{tikzpicture}
  \end{center}
  With powers defined recursively:
  \begin{align*}
    L^0 &:= \epsilon\\
    L^1 &:= L\\
    L^{i+1} &:= \sum_{w \in \Sigma^*} ((L^i,u)\cdot (L,v))w, 
    \hspace{0.5cm} \text{where} \hspace{0.5cm} w = uv.
  \end{align*}
  Kleene Star can now also be defined for formal power series:
  \[
    L^* := L^0 + L^1 + L^2 + L^3 + \dots = \sum_{i \in \mathbb{N}_0} L^i.
  \]
\end{definition}
With these operation defined our two versions of formal languages are 
powerful tools, which will be compared in the following sections. In order
to conpare them we will also need finite automatas. Simillar to formal languages
there are multiple ways one can define finite automatas. In this paper we
will define two. One is more suitable for combining it with the set version
of formal languages and the other one is more siutable for the formal power
series representation. First we will define the one more suitable for the
set version.
\begin{definition}
  A finite automata is a quintuple $M = (Q, \Sigma, \delta, q_0, F)$, 
  where
  \begin{enumerate}
    \item Q is a finite set of states,
    \item $\Sigma$ is an alphabet, all inputs are constructed from,
    \item $\delta: Q \times \Sigma \to Q$ is the transition function,
      carrying the automata from one state to another,
    \item $q_0 \in Q$ is the initial state of the automata,
    \item $F \subset Q$ are the accepting states of the automata.
  \end{enumerate}
  %Definition is out of theoretical computer science book (german)
\end{definition}
\begin{example}
  Here is a simple automata depicted, where the circles are states
  , the arrows are transitions, the start is indicating the initial 
  state, and the double lined circle is the accepting state.
  $M = ()$
    \begin{tikzpicture}
      \node[state,initial] (q1) at (0,0) {$s_1$};
      \node[state,accepting] (q2) at (3,0) {$s_2$};
      \draw (q1) edge[loop above] node{x} (q1)
      (q2) edge[loop above] node{z} (q2);
      \draw[->](0.5,0)--(2.5,0);
      \node[fill=white] at (1.5,0.3) {y};
    \end{tikzpicture}
\end{example}

% Define Formal languages (Sets and Power sereies) 
% and automatas (Tuple and Matrix)
\section{Formal Languages: Sets vs. Formal Power Series}
% translate one in another
% Formal power series are an algebra over Boolean algebra
% U could define the same algebra for sets with set operations
% But u could replace the boolean algebra by any abritary ring
% Which makes formal power series more powerful cuz u can abstract even more
\section{Maximum Gap Lemma}
\section{Pumping Lemma}
\section{Not all Formal languages are regular}
\section{Pumping Lemma vs. Maximum Gap Lemma}
\section{Kleene's theorem (Power series, Matrix Automata)}
\section{Kleene's theorem (Set, Tuple)}
\section{Kleene's theorem vs. Kleene's theorem}
\section{Kleene Schützenberg theorem and weigthed automatas}
\section{Conclusion}

>>>>

we can always number a finite amount of objects. The above definition

@@

  $\Sigma$. The empty word which contains no characters, 

@@

M is a set of words and multiplication is concetanation of words, then 

@@

  and
  \[ w^0 = \epsilon.\]

@@

    \end{tikzpicture}
  \end{center}
  With powers defined recursively:
    L^0 &:= \epsilon\\
    L^1 &:= L\\
    L^{i+1} &:= \sum_{w \in \Sigma^*} ((L^i,u)\cdot (L,v))w, 
    \hspace{0.5cm} \text{where} \hspace{0.5cm} w = uv.
  Kleene Star can now also be defined for formal power series:
    L^* := L^0 + L^1 + L^2 + L^3 + \dots = \sum_{i \in \mathbb{N}_0} L^i.
With these operation defined our two versions of formal languages are 
powerful tools, which will be compared in the following sections. In order
to conpare them we will also need finite automatas. Simillar to formal languages
there are multiple ways one can define finite automatas. In this paper we
will define two. One is more suitable for combining it with the set version
of formal languages and the other one is more siutable for the formal power
series representation. First we will define the one more suitable for the
set version.
  A finite automata is a quintuple $M = (Q, \Sigma, \delta, q_0, F)$, 
  where
    \item Q is a finite set of states,
    \item $\Sigma$ is an alphabet, all inputs are constructed from,
    \item $\delta: Q \times \Sigma \to Q$ is the transition function,
      carrying the automata from one state to another,
    \item $q_0 \in Q$ is the initial state of the automata,
    \item $F \subset Q$ are the accepting states of the automata.
  %Definition is out of theoretical computer science book (german)
In simple terms, an finite automata gets an input string and a starting 
state. Then it works through the input string, starting at the first charater,
carrying the starting state to another according to the transition function.
Exactly the same procedure, for the second input string character and the 
remainding characters. If this process ends up in an accepting state, 
the automata accepts the input string. Consequently, 
the finite automata generates a language, namely the words that the automata 
accepts. The definition above is one possible way to encode a finite automata. 
Like with graphs we can encode the transition function of an finite
automata, with just a matrix.
  Let $M = (Q, \Sigma, \delta, q_0, F)$ be an abitrary finite 
  automata, where $Q = \{ s_1, s_2, \dots s_n\}.$ \\
  Frthermore, we need $H(i,j) = \{ \alpha \in \Sigma : \delta(s_i, \alpha) = s_j\}$
  and two helper functions:
    w: \Sigma^\times \to \Sigma^*, 
    \delta': \mathbb{N} \times \mathbb{N} \to \Sigma^*
  where
    w(\{a_1,a_2,\dots, a_n\}) = a_1a_2\dots a_n
    \delta'(i, j) = 
      \begin{cases}
        w(H(i,j)), & \text{if} \hspace{0.2cm} \exists \alpha \in \Sigma: 
          \delta(s_i, \alpha) = s_j \\
        0, & \text{otherwise}
      \end{cases}.
  Then the finite automata $M$ can be encoded with a Matrix
    $A \in (\Sigma^*)^{n \times n}$
  in the following way:
    A = 
    \begin{pmatrix}
      \delta'(1,1)       & \delta'(1,2) & \delta'(1,3) & \dots & \delta'(1,n) \\
      \delta'(2,1)       & \delta'(2,2) & \delta'(2,3) & \dots & \delta'(2,n) \\
      \vdots             & \vdots       & \vdots       & \ddots& \vdots\\
      \delta'(n,1)       & \delta'(n,2) & \delta'(n,3) & \dots & \delta'(n,n) \\
    \end{pmatrix}
  .\]
  We can denote a state transition $s_i \to s_j$ with $s_iA_{ij} = s_j$.
  The kleene star of a Matrix can intuively  be defined as:
    A^* = E_n + A + A^2 + A^3 \dots  = \sum_{i \in \mathbb{N}_0} A^i.
  Now we can write our finite automata in a shorter way:
    M' = (\Sigma, A, q_0, F)
This definition only works, because we have finite sets. Note that we can omit
the state set, since as mentioned above, there always exist an isomorphism
between a set with finite objects and a subset of the natural numbers. The first
object gets the number one, the second the number two, and so forth. The new
definition of the state transition function works like a lookup table, when one
wants to know what characters map state $i$ to state $j$, its a simple lookup:
$M_{ij}$.

\begin{example}
  Here is a simple automata depicted, where the circles are states
  , the arrows are transitions, the start is indicating the initial 
  state, and the double lined circle is the accepting state.
  \[ 
    M_\text{tuple} = (\{s_1, s_2\}, \{a,b,c\}, \delta_0, s_1, \{s_2\}),
    M_\text{matrix} = (\{a,b,c\}, A, s_1, \{ s_2 \})
  \]\\
  where
  \[
    \delta_0(s_1, a) = s_1, \hspace{0.2cm} 
    \delta_0(s_1, b) = s_2, \hspace{0.2cm} \delta_0(s_2,c) = s_2
  \]
  and
  \[ 
    A = 
    \begin{pmatrix}
      a & b \\
      0 & c 
    \end{pmatrix}.
  \]
  \begin{center}
    \begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
      \node[state,initial] (q1) at (0,0) {$s_1$};
      \node[state,accepting] (q2) at (3,0) {$s_2$};
      \path[->] (q1) edge[loop above] node{a} (q1)
      (q2) edge[loop above] node{c} (q2)
      (q1) edge[bend left] node{b} (q2);
    \end{tikzpicture}
  \end{center}
  The resulting accepting formal language is:
  \[ \text{Set represntation: } L = \{a^nbc^n : n \in \mathbb{N}\}\]
  \[
    \text{Formal power series represntation: } L = a^*\cdot b\cdot c^* 
  = a^*bc^*
\]
\end{example}
% Define Formal languages (Sets and Power sereies) 
% and automatas (Tuple and Matrix)
\section{Formal Languages: Sets vs. Formal Power Series}
% translate one in another
% Formal power series are an algebra over Boolean algebra
% U could define the same algebra for sets with set operations
% But u could replace the boolean algebra by any abritary ring
% Which makes formal power series more powerful cuz u can abstract even more
\section{Maximum Gap Lemma}
\section{Pumping Lemma}
\section{Not all Formal languages are regular}
\section{Pumping Lemma vs. Maximum Gap Lemma}
\section{Kleene's theorem (Power series, Matrix Automata)}
\section{Kleene's theorem (Set, Tuple)}
\section{Kleene's theorem vs. Kleene's theorem}
\section{Kleene Schützenberg theorem and weigthed automatas}
\section{Conclusion}
